---
title: "GA convergence test"
author: "Charles Whorton"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


if (!require(tidyverse)){
  install.packages('tidyverse')
}
library(tidyverse)

if (!require(DT)){
  install.packages('DT')
}
library(DT)

if (!require(htmlwidgets)){
  install.packages('htmlwidgets')
}
library(htmlwidgets)

if (!require(readxl)){
  install.packages('readxl')
}
library(readxl)

if (!require(GA)){
  install.packages('GA')
}
library(GA)

if (!require(httr)){
  install.packages('httr')
}
library(httr)

if (!require(jsonlite)){
  install.packages('jsonlite')
}
library(jsonlite)

if (!require(ggplot2)){
  install.packages('ggplot2')
}
library(ggplot2)

```

```{r}

### Import and Clean Data ###
temp_file = "/Users/charleswhorton/Desktop/GradSchool/Capstone/capstone/Data/2022_MLS_GoalsAdded_and_Salary_Data.xlsx"

Player_Goals_Added <- read_excel(temp_file, sheet = "Player_Goals_Added")
Player_Salary = read_excel(temp_file, sheet = "Player_Salary") %>%
  select(-c("Season"))

MLSPA_Player_Salary_202209 =
  read_csv("http://s3.amazonaws.com/mlspa/9_2_2022-Roster-Freeze-Salary-List.csv?mtime=20221017131703") %>%
  select(-c("Nickname")) %>%
  mutate(`2022 Base Salary` = floor(as.numeric(gsub("[$,]", "", `2022 Base Salary`)))) %>%
  mutate(`2022 Guar. Comp.` = floor(as.numeric(gsub("[$,]", "", `2022 Guar. Comp.`)))) %>%
  mutate(Source = 9)

MLSPA_Player_Salary202204 = 
  read_csv("http://s3.amazonaws.com/mlspa/2022_salary_list_4.15__for_site.csv?mtime=20220517164257") %>%
  mutate(`2022 Base Salary` = floor(as.numeric(gsub("[$,]", "", `2022 Base Salary`)))) %>%
  mutate(`2022 Guar. Comp.` = floor(as.numeric(gsub("[$,]", "", `2022 Guar. Comp.`)))) %>%
  mutate(Source = 4)

MLSPA_Player_Salary202204 = rename(MLSPA_Player_Salary202204, "Position" = "Playing Position")

MLSPA_Player_Salary =
  union(MLSPA_Player_Salary_202209, MLSPA_Player_Salary202204)
  # MLSPA_Player_Salary202204

MLSPA_Player_Salary = MLSPA_Player_Salary %>%
  mutate(Player = paste(MLSPA_Player_Salary$'First Name', MLSPA_Player_Salary$'Last Name', sep = " "))

MLSPA_Player_Salary = MLSPA_Player_Salary %>%
  relocate(c("Player", "Club", "Position", "2022 Base Salary", "2022 Guar. Comp."))

MLSPA_Player_Salary = rename(MLSPA_Player_Salary,
                             "Team" = "Club",
                             "Base Salary" = "2022 Base Salary",
                             "Guaranteed Compensation" = "2022 Guar. Comp.") %>%
  select(-c("Last Name", "First Name", Team, Position))

MLSPA_Player_Salary = MLSPA_Player_Salary %>%
  arrange(desc(Player)) %>%
  group_by(Player) %>%
  mutate(Rank = row_number(desc(Source))) %>%
  filter(Rank == 1)

Salary_Data =
  full_join(Player_Salary, MLSPA_Player_Salary %>% select(-c(Source, Rank)), by = 'Player')

Player_Data = Player_Goals_Added %>% 
  left_join(Salary_Data, by = "Player")

Player_Data = Player_Data %>% 
  mutate('Guaranteed Compensation.x' 
         = coalesce(Player_Data$'Guaranteed Compensation.x',Player_Data$'Guaranteed Compensation.y')) %>%
  mutate('Base Salary.x' 
         = coalesce(Player_Data$'Base Salary.x',Player_Data$'Base Salary.y')) %>%
  mutate('Position.x' 
         = coalesce(Player_Data$'Position.x',Player_Data$'Position.y')) %>%
  mutate('Team.x' 
         = coalesce(Player_Data$'Team.x',Player_Data$'Team.y')) %>%
  select(-c("Team.y",
            "Position.y",
            "Base Salary.y", 
            "Guaranteed Compensation.y")) %>% 
  rename("Position" = "Position.x",
         "Team" = "Team.x",
         "Base_Salary" = "Base Salary.x",
         "Guaranteed_Compensation" = "Guaranteed Compensation.x",
         "Goals_Added" = "Goals Added") %>% 
  drop_na(Base_Salary)

Player_Data %>%
  filter(is.na(Base_Salary))

Player_Data <- data.frame(Player_Data %>% 
  mutate(Log_Goals_Added = log10(Goals_Added+(abs(min(Player_Data$Goals_Added))+1.01))))

```

```{r}
  
  ### Initialize population
      initializePopulation = function(k){
        
        function(GA){
          mat = matrix(0, ncol = GA@nBits, nrow = GA@popSize)
          
          for(i in seq_len(GA@popSize))
            mat[i, sample(GA@nBits, k)] = 1
          
          for (i in 1:length(position_values$input_value)) {
            if (position_values$input_value[i] > 0) {
              ind = which(Player_Data$Position 
                          == names(position_values$input_value)[i])
              mat[, ind[sample(length(ind), 1)]] = 1
            }
          }
          
          mat
        }
      }
      
      ### Crossover function
        # custom function that performs crossover while keeping the number of 
        # selected bits equal to the selected population
      crossoverFunction = function(GA, parents){
        
        parents = GA@population[parents,] %>%
          apply(1, function(x) which(x == 1)) %>%
          t()
        
        parents_diff = list("vector", 2)
        parents_diff[[1]] = setdiff(parents[2,], parents[1,])
        parents_diff[[2]] = setdiff(parents[1,], parents[2,])
        
        children_ind = list("vector", 2)
        for(i in 1:2){
          k = length(parents_diff[[i]])
          change_k = sample(k, sample(ceiling(k/2), 1))
          children_ind[[i]] = if(length(change_k) > 0){
            c(parents[i, -change_k], parents_diff[[i]][change_k])
          } else {
            parents[i,]
          }
        }
        
        children = matrix(0, nrow = 2, ncol = GA@nBits)
        for(i in 1:2)
          children[i, children_ind[[i]]] = 1
        
        list(children = children, fitness = c(NA, NA))
      }
      
      ### Mutation function
        # custom function that performs mutation while keeping the number of 
        # selected bits equal to the selected population
      mutationFunction = function(GA, parent){
        ind = which(GA@population[parent,] == 1)
        n_change = sample(sel_pop, 1)
        ind[sample(length(ind), n_change)] = 
          sample(setdiff(seq_len(GA@nBits), ind), n_change)
        parent = integer(GA@nBits)
        parent[ind] = 1
        
        parent
      }
      
      ### Fitness function
      # objective function that sums the goals_added value and checks the
      # constraint of number of players and their positions
      fitness = function(x)
      {
        current_goals_added=x%*%Player_Data$Log_Goals_Added
        current_salary=x%*%Player_Data$Guaranteed_Compensation
        positions = Player_Data$Position[x == 1] # positions of selected players
        positions_count = table(positions) # count of players in each position
        
        if(current_salary>max_salary) {
          return(0) # Penalty for exceeding the max salary
        }
        else if (all(positions_count == positions_required)) {
          return(current_goals_added) # meets the players and positions constraint
        }
        else {
          return(0)
        }
      }
```

```{r}

  n = 200
  sel_pop = 5
  max_salary = 1750000
  
  # n = 200
  # sel_pop = 5
  # max_salary = 1750000
  
  position_values = data.frame(position = c("ST","CM","CB"),
                                   input_value = c(2,2,1))
  
  positions_required = position_values %>% 
        spread(position, input_value, fill = 0)
  
  # Original genetic algorithm function
      # GA = ga(
      #   type = "binary",
      #   fitness = fitness,
      #   nBits = nrow(Filtered_Player_Data),
      #   population = initializePopulation(sel_pop),
      #   crossover = crossoverFunction,
      #   mutation = mutationFunction,
      #   run = 125,
      #   pmutation = .25,
      #   maxiter = 300,
      #   popSize = 30,
      #   monitor = FALSE
      # )

```


# Testing the highest fitness value per population size

```{r}
  # 190
  pop_test_data = data.frame(popSize = numeric(0), 
                         value = numeric(0))
  
  pop_size = c(seq(10, 200, by = 10))

  for (p in pop_size){
    fitnessValues = c()
    for (x in 1:10) {
      GA <- ga(
        type = "binary",
        fitness = fitness,
        nBits = nrow(Player_Data),
        population = initializePopulation(sel_pop),
        crossover = crossoverFunction,
        mutation = mutationFunction,
        run = 225,
        pmutation = .7,
        maxiter = 950,
        popSize = p,
        monitor = FALSE
      )
      
      fitnessValues = append(fitnessValues, GA@fitnessValue)
    }
  
    pop_test_data = pop_test_data %>% 
      add_row(popSize = p, 
              value = mean(fitnessValues))
  }    

pop_test_data %>% filter(value == max(pop_test_data$value))
  
pop_test_data
```

# Plotting the highest fitness value per population size

```{r}
# Create line graph with multiple lines
ggplot(data = pop_test_data, 
       aes(x = popSize, 
           y = value)) +
  geom_smooth(method = "loess", span = 0.35, se = FALSE) +
  geom_line() +
  geom_vline(xintercept = sort(subset(pop_test_data, 
                                      value == max(pop_test_data$value))[[1]])[1],
             colour="red") +
  scale_x_continuous(breaks = seq(10, 200, by = 20)) +
  labs(x = "Population Size",
       y = "Fitness Value",
       title = "Figure 3",
       subtitle = "Objective function fitness value by population size") +
  theme_bw()

```

# Testing the highest fitness value per runs

```{r}
  #225
  run_test_data = data.frame(run_nums = numeric(0), 
                         value = numeric(0))
  
  run_nums = c(seq(50, 500, by = 25))

  for (r in run_nums){
    fitnessValues = c()
    for (x in 1:10) {
      GA <- ga(
        type = "binary",
        fitness = fitness,
        nBits = nrow(Player_Data),
        population = initializePopulation(sel_pop),
        crossover = crossoverFunction,
        mutation = mutationFunction,
        run = r,
        pmutation = .7,
        maxiter = 950,
        popSize = 200,
        monitor = FALSE
      )
      
      fitnessValues = append(fitnessValues, GA@fitnessValue)
    }
  
    run_test_data = run_test_data %>% 
      add_row(run_nums = r, 
              value = mean(fitnessValues))
  }    

run_test_data %>% filter(value == max(run_test_data$value))
  
run_test_data
```

# Plotting the highest fitness value per runs

```{r}
# Create line graph with multiple lines
ggplot(data = run_test_data, 
       aes(x = run_nums, 
           y = value)) +
  geom_smooth(method = "loess", span = 0.5, se = FALSE) +
  geom_line() +
  geom_vline(xintercept = sort(subset(run_test_data, 
                                      value == max(run_test_data$value))[[1]])[1],
             colour="red") +
  scale_x_continuous(breaks = seq(50, 500, by = 50)) +

  labs(x = "Number of Runs",
       y = "Fitness Value",
       title = "Figure 2",
       subtitle = "Objective function fitness value by number of runs") +
  theme_bw()

```

# Testing the "max_iter" parameter

```{r}
#950
max_iter_test_data = data.frame(max_iterations = numeric(0), 
                       value = numeric(0))

max_iterations = c(seq(100, 1000, by = 25))

for (m in max_iterations) {
    fitnessValues = c()
    for (x in 1:10) {
      GA <- ga(
        type = "binary",
        fitness = fitness,
        nBits = nrow(Player_Data),
        population = initializePopulation(sel_pop),
        crossover = crossoverFunction,
        mutation = mutationFunction,
        run = 225,
        pmutation = .7,
        maxiter = m,
        popSize = 200,
        monitor = FALSE
      )
      
      fitnessValues = append(fitnessValues, GA@fitnessValue)
    }
  
          max_iter_test_data = max_iter_test_data %>% 
      add_row(max_iterations = m,
              value = mean(fitnessValues))
}

max_iter_test_data %>% filter(value == max(max_iter_test_data$value))

```

# Plotting the highest fitness value per max iteration value

```{r}
# Create line graph with multiple lines
ggplot(data = max_iter_test_data, 
       aes(x = max_iterations, 
           y = value, 
           # color = factor(attempt_num), 
           # linetype = factor(attempt_num)
           )) +
  geom_line() +
  geom_smooth(method = "loess", span = 0.5, se = FALSE) +
  geom_vline(xintercept = sort(subset(max_iter_test_data, 
                                      value == max(max_iter_test_data$value))[[1]])[1],
             colour="red") +
  scale_x_continuous(breaks = seq(100, 1000, by = 100)) +
  # geom_bar(stat = "identity") +
  labs(x = "Max number of iterations",
       y = "Fitness Value",
       title = "Figure 1", 
       subtitle = "Objective function fitness value by max number of iterations") +
  theme_bw()


```

# Testing the highest fitness value per mutation probability value

```{r}

mutation_test_data = data.frame(p_mutation = numeric(0), 
                         value = numeric(0))
    
  p_mutation = c(seq(.1, .95, by = .05))
  
for (p in p_mutation) {
    fitnessValues = c()
    for (x in 1:10) {
      GA <- ga(
        type = "binary",
        fitness = fitness,
        nBits = nrow(Player_Data),
        population = initializePopulation(sel_pop),
        crossover = crossoverFunction,
        mutation = mutationFunction,
        run = 225,
        pmutation = p,
        maxiter = 950,
        popSize = 200,
        monitor = FALSE
      )
      
      fitnessValues = append(fitnessValues, GA@fitnessValue)
    }
  
          mutation_test_data = mutation_test_data %>% 
      add_row(p_mutation = p,
              value = mean(fitnessValues))
}

mutation_test_data

subset(mutation_test_data, value == max(mutation_test_data$value))[[1]]

```

# Plotting the highest fitness value per mutation probability value

```{r}

ggplot(data = mutation_test_data, 
       aes(x = p_mutation, y = value)) + 
  geom_smooth(method = "loess", span = 0.65, se = FALSE) +
  geom_line() +
  geom_vline(xintercept = sort(subset(mutation_test_data, 
                                      value == max(mutation_test_data$value))[[1]])[1],
             colour="red") +
  scale_x_continuous(breaks = seq(.1, .9, by = .1)) +
  labs(x = "Mutation Probability", 
       y = "Fitness Value", 
       title = "Figure 5", 
       subtitle = "Objective function fitness value by mutation probability") + 
  theme_bw()

```

# Testing the algorithm execution time

```{r}

time_test_data = data.frame(iteration = numeric(0), 
                         time = numeric(0))
# pm = sort(subset(mutation_test_data, value == max(mutation_test_data$value))[[1]])[1]
  
    for (x in 1:50) {
      execution_time <- system.time(
      GA <- ga(
        type = "binary",
        fitness = fitness,
        nBits = nrow(Player_Data),
        population = myInit(sel_pop),
        crossover = myCrossover,
        mutation = myMutation,
        run = 225,
        pmutation = .7,
        maxiter = 950,
        popSize = 200,
        monitor = FALSE
      ))[[3]]
      
      time_test_data = time_test_data %>% 
      add_row(iteration = x,
              time = execution_time)
    }
  
mean(time_test_data$time)

```

```{r}

ggplot(Player_Data, aes(x=Guaranteed_Compensation, y=Log_Goals_Added)) + 
  geom_point() + 
  geom_smooth(method=lm)

```

# Testing the algorithm execution time

```{r}

time_test_data = data.frame(iteration = numeric(0), 
                         time = numeric(0))
  
    for (x in 1:50) {
      execution_time <- system.time(
      GA <- ga(
        type = "binary",
        fitness = fitness,
        nBits = nrow(Player_Data),
        population = myInit(sel_pop),
        crossover = myCrossover,
        mutation = myMutation,
        run = 125,
        pmutation = .25,
        maxiter = 300,
        popSize = 30,
        monitor = FALSE
      ))[[3]]
      
      time_test_data = time_test_data %>% 
      add_row(iteration = x,
              time = execution_time)
    }
  
mean(time_test_data$time)

```

```{r}

ggplot(Player_Data, aes(x=Guaranteed_Compensation, y=Log_Goals_Added)) + 
  geom_point() + 
  geom_smooth(method=lm)

```


```{r}
test_data = data.frame(exec_num = numeric(0), 
                         value = numeric(0))

  
    fitnessValues = c()
    for (x in 1:10) {
      GA <- ga(
        type = "binary",
        fitness = fitness,
        nBits = nrow(Player_Data),
        population = initializePopulation(sel_pop),
        crossover = crossoverFunction,
        mutation = mutationFunction,
        run = 225,
        pmutation = .7,
        maxiter = 950,
        popSize = 200,
        monitor = FALSE
      )
      
      fitnessValues = append(fitnessValues, GA@fitnessValue)
    }
  
test_data = test_data %>% 
  add_row(exec_num = c(1:10), 
          value = mean(fitnessValues))
# }

mean(test_data$value)

```


```{r}

test_data = data.frame(exec_num = numeric(0), 
                         value = numeric(0))

  
    fitnessValues = c()
    for (x in 1:1) {
      GA <- ga(
        type = "binary",
        fitness = fitness,
        nBits = nrow(Player_Data),
        population = initializePopulation(sel_pop),
        crossover = crossoverFunction,
        mutation = mutationFunction,
        run = 125,
        pmutation = .25,
        maxiter = 300,
        popSize = 30,
        monitor = FALSE
      )
      
      fitnessValues = append(fitnessValues, GA@fitnessValue)
    }
  
test_data = test_data %>% 
  add_row(exec_num = c(1:10), 
          value = mean(fitnessValues))
# }

mean(test_data$value)

```